Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2025-07-22 index.tex
fix reference

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{interact}

\usepackage{amsmath,amssymb}
\usepackage{setspace}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{5}
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks},
  pdfauthor={H. Sherry Zhang; Roger D. Peng},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While some researchers
may struggle to identify the root causes, experienced researchers can
often quickly diagnose problems by checking a few key assumptions. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
introduce the term \emph{analysis validation checks} to formalize and
externalize these informal assumptions. We then introduce a procedure to
identify a subset of checks that best predict the occurrence of
unexpected outcomes, based on simulations of the original data. The
checks are evaluated in terms of accuracy, determined by binary
classification metrics, and independence, which measures the shared
information among checks. We demonstrate this approach with a toy
example using step count data and a generalized linear model example
examining the effect of particulate matter air pollution on daily
mortality.
\end{abstract}

@@

\begin{verbatim}
Orientation inferred to be along y-axis; override with
`position_quasirandom(orientation = 'x')`
\end{verbatim}

@@

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k)\]

@@

The typical approach to studying the association between PM10 and
mortality is to apply a generalized linear model with a Poisson link to
relate daily mortality counts to daily measures of PM10. Based on
previous work and the range of effect sizes published in the literature,
an analyst might expect the coefficient for PM10 in this GLM to lie
between \([0, 0.005]\), after adjusting for daily temperature
\citep{samet2000fine, welty2005acute}. Note that the relatively simple
modeling approach being used here is primarily for demonstration;
typically far more sophisticated semi-parametric approaches are used in
the literature \citep{peng2006model}.

@@

\begin{verbatim}
Warning: There were 990 warnings in `mutate()`.
The first warning was:
i In argument: `fit = list(...)`.
i In row 2.
Caused by warning in `dpois()`:
! non-integer x = 291.790133
i Run `dplyr::last_dplyr_warnings()` to see the 989 remaining warnings.
\end{verbatim}

\begin{verbatim}
Joining with `by = join_by(unexpect)`
\end{verbatim}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
\toprule
Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
\midrule
Check 1: cor(m, PM10) < -0.03 & 0.518 & 0.888 & 1.000 & 0.740 & 0.802\\
Check 2: cor(m, tmp) > -0.35 & 0.259 & 0.543 & 1.000 & 0.448 & 0.601\\
Check 3: mortality outlier & 0.281 & 0.833 & 1.000 & 0.521 & 0.705\\
Check 4: PM10 outlier & 0.289 & 0.871 & 1.000 & 0.535 & 0.720\\
Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.874 & 0.667 & 0.779 & 0.764 & 0.773\\
\bottomrule
\end{tabular}}

@@

\begin{quote}
(cor(mortality, PM10) \textless{} -0.03) AND (cor(mortality, tmp)
\textgreater{} -0.3 OR (mortality outlier AND PM10 outlier))
\end{quote}

This generates a precision of 0.874 and a recall of 0.667 for predicting
the unexpected PM10 coefficient. As shown in
Figure~\ref{fig-linear-reg-tree}, there is no single analysis check in
the tree that predicts an unexpected outcome. rather at least three
checks in the tree must be TRUE in order for the model to predict an
unexpected outcome. Given the high independence of the checks
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.

@@

\begin{verbatim}
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 3288 Columns: 6
-- Column specification --------------------------------------------------------
Delimiter: ","
chr  (1): season
dbl  (4): mortality, pm10, temp, dewpt
date (1): date

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Rows: 40 Columns: 2
-- Column specification --------------------------------------------------------
Delimiter: ","
chr (2): city, cityname

i Use `spec()` to retrieve the full column specification for this data.
i Specify the column types or set `show_col_types = FALSE` to quiet this message.
Joining with `by = join_by(city)`
\end{verbatim}

>>>>

% Options for packages loaded elsewhere
% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
  12pt,
]{interact}
\usepackage{xcolor}
\usepackage{amsmath,amssymb}
\setcounter{secnumdepth}{5}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math} % this also loads fontspec
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else
  % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{setspace}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
% Make \paragraph and \subparagraph free-standing
\makeatletter
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}{
    \@ifstar
      \xxxParagraphStar
      \xxxParagraphNoStar
  }
  \newcommand{\xxxParagraphStar}[1]{\oldparagraph*{#1}\mbox{}}
  \newcommand{\xxxParagraphNoStar}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}{
    \@ifstar
      \xxxSubParagraphStar
      \xxxSubParagraphNoStar
  }
  \newcommand{\xxxSubParagraphStar}[1]{\oldsubparagraph*{#1}\mbox{}}
  \newcommand{\xxxSubParagraphNoStar}[1]{\oldsubparagraph{#1}\mbox{}}
\fi
\makeatother

\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\newsavebox\pandoc@box
\newcommand*\pandocbounded[1]{% scales image to fit in text height/width
  \sbox\pandoc@box{#1}%
  \Gscale@div\@tempa{\textheight}{\dimexpr\ht\pandoc@box+\dp\pandoc@box\relax}%
  \Gscale@div\@tempb{\linewidth}{\wd\pandoc@box}%
  \ifdim\@tempb\p@<\@tempa\p@\let\@tempa\@tempb\fi% select the smaller of both
  \ifdim\@tempa\p@<\p@\scalebox{\@tempa}{\usebox\pandoc@box}%
  \else\usebox{\pandoc@box}%
  \fi%
}
% Set default figure placement to htbp
\def\fps@figure{htbp}
\makeatother

\setlength{\emergencystretch}{3em} % prevent overfull lines

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[]{natbib}
\bibliographystyle{plainnat}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{orcidlink}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\usepackage{bookmark}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same}
\hypersetup{
  pdftitle={Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks},
  pdfauthor={H. Sherry Zhang; Roger D. Peng},
  pdfkeywords={diagnostic, logic regression, data analysis assumptions},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While some researchers
may struggle to identify the root causes, experienced researchers can
often quickly diagnose problems by checking a few key assumptions. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
introduce the term \emph{analysis validation checks} to formalize and
externalize these informal assumptions. We then introduce a procedure to
identify a subset of checks that best predict the occurrence of
unexpected outcomes, based on simulations of the original data. The
checks are evaluated in terms of accuracy, determined by binary
classification metrics, and independence, which measures the shared
information among checks. We demonstrate this approach with a toy
example using step count data and a generalized linear model example
examining the effect of particulate matter air pollution on daily
mortality.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
diagnostic, logic regression, data analysis assumptions\sep 
diagnostic, logic regression, data analysis assumptions
\end{keywords}

@@

@@

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_k),\]

@@

The typical approach to studying the association between PM10 and
mortality is to apply a generalized linear model (Poisson family with a
log link) to relate daily mortality counts to daily measures of PM10.
Based on previous work and the range of effect sizes published in the
literature, an analyst might expect the coefficient for PM10 in this GLM
to lie between \([0, 0.005]\), after adjusting for daily temperature
\citep{samet2000fine, welty2005acute}. Note that the relatively simple
modeling approach being used here is primarily for demonstration;
typically far more sophisticated semi-parametric approaches are used in
the literature \citep{peng2006model}.

@@

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
\toprule
Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
\midrule
Check 1: cor(m, PM10) < -- 0.03 & 0.551 & 0.894 & 1.000 & 0.763 & 0.815\\
Check 2: cor(m, tmp) > -- 0.35 & 0.281 & 0.556 & 1.000 & 0.472 & 0.612\\
Check 3: mortality outlier & 0.302 & 0.775 & 1.000 & 0.536 & 0.692\\
Check 4: PM10 outlier & 0.305 & 0.766 & 1.000 & 0.537 & 0.690\\
Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.876 & 0.722 & 0.812 & 0.798 & 0.803\\
\bottomrule
\end{tabular}}

@@

\begin{quote}
(cor(mortality, PM10) \textless{} -- 0.03) AND (cor(mortality, tmp)
\textgreater{} -- 0.3 OR (mortality outlier AND PM10 outlier))
\end{quote}

This generates a precision of 0.876 and a recall of 0.722 for predicting
the unexpected PM10 coefficient. As shown in
Figure~\ref{fig-linear-reg-tree}, there is no single analysis check in
the tree that predicts an unexpected outcome. rather at least three
checks in the tree must be TRUE in order for the model to predict an
unexpected outcome. Given the high independence of the checks
(Table~\ref{tbl-linear-reg}), this suggests that unexpected results are
only likely after multiple anomalies are observed in the data.

@@
